---
title: "Homework 5" 
author: "Yiying Wu (yw3996)"
output:
  pdf_document:
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[L]{Data Science 2 HW5}
- \fancyhead[R]{Yiying Wu (yw3996)}
- \fancypagestyle{plain}{\pagestyle{fancy}}
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## R packages
```{r}
library(tidyverse)
library(caret)
library(tidymodels)
library(ISLR)
library(e1071)
library(factoextra)
```

## 1. auto.csv data

In this problem, we will apply support vector machines to predict whether a given car gets high or low gas mileage based on the dataset “auto.csv” (used in Homework 3; see Homework 3 for more details of the dataset). The response variable is mpg cat. The predictors are cylinders, displacement, horsepower, weight, acceleration, year, and origin. Split the dataset into two parts: training data (70%) and test data (30%).


Input dataset
```{r}
dat<-read_csv("./data/auto.csv")%>%
  mutate(
    mpg_cat = as.factor(mpg_cat),
    origin = as.factor(origin))
dat <- dat%>%
  na.omit()
```

**Response: mpg_cat**
```{r}
contrasts(dat$mpg_cat)
```


Split the dataset into two parts: training data (70%) and test data (30%).

```{r}
set.seed(1)
data_split <- initial_split(dat, prop = 0.7)

# Extract the training and test data
training_data <- training(data_split)
testing_data <- testing(data_split)

ctrl <- trainControl(method = "cv", number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```


### (a) Fit a support vector classifier to the training data. What are the training and test error rates?
```{r}
set.seed(1)
linear.tune <- tune.svm(mpg_cat ~ . , 
                        data = training_data, 
                        kernel = "linear", 
                        cost = exp(seq(-5,2, len = 50)),
                        scale = TRUE)
plot(linear.tune)
# summary(linear.tune)
linear.tune$best.parameters

best.linear <- linear.tune$best.model
summary(best.linear)

pred.linear <- predict(best.linear, newdata = testing_data)

confusionMatrix(data = pred.linear, 
                reference = testing_data$mpg_cat)

```

The **training error rate** is 0.0119.

Test Error Rate $= 1 - Accuracy=1-0.9068=0.0932$

The **test error rate** for the model on the testing data is approximately 0.0932.

### (b) Fit a support vector machine with a radial kernel to the training data. What are the training and test error rates?
```{r}
set.seed(1)
radial.tune <- tune.svm(mpg_cat ~ . , 
                        data = training_data, 
                        kernel = "radial", 
                        cost = exp(seq(1, 7, len = 50)),
                        gamma = exp(seq(-10, -2,len = 20)))

plot(radial.tune, transform.y = log, transform.x = log, 
     color.palette = terrain.colors)
# summary(radial.tune)

radial.tune$best.parameters

best.radial <- radial.tune$best.model
summary(best.radial)

# Predict on the training data using the best model
pred.radial.train <- predict(best.radial, newdata = training_data)

# Calculate the confusion matrix for the training predictions
conf.matrix.train <- confusionMatrix(data = pred.radial.train, 
                                     reference = training_data$mpg_cat)

# Extract and print the training error rate
train.error.rate <- 1 - conf.matrix.train$overall['Accuracy']
print(train.error.rate)

pred.radial <- predict(best.radial, newdata = testing_data)

confusionMatrix(data = pred.radial, 
                reference = testing_data$mpg_cat)
```

The **training error rate** is `r round(train.error.rate,4)`.

Test Error Rate $= 1 - Accuracy=1-0.9153=0.0847$

The **test error rate** for the model on the testing data is approximately 0.0847.

## 2. USArrests data

In this problem, we perform hierarchical clustering on the states using the USArrests data in the ISLR package. For each of the 50 states in the United States, the dataset contains the number of arrests per 100,000 residents for each of three crimes: Assault, Murder, and Rape. The dataset also contains the percent of the population in each state living in urban areas, UrbanPop. The four variables will be used as features for clustering.

```{r}
data(USArrests)
dat2 <- na.omit(USArrests)
set.seed(1)
```


### (a) Using hierarchical clustering with complete linkage and Euclidean distance, cluster the states. Cut the dendrogram at a height that results in three distinct clusters. Which states belong to which clusters?

Use hierarchical clustering with complete linkage and Euclidean distance, cluster the states

```{r}
hc.complete <- hclust(dist(dat2), method = "complete")
```

Visualize the dendrogram and cut the dendrogram at a height that results in three distinct clusters

```{r}
fviz_dend(hc.complete, k = 3,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)

ind4.complete <- cutree(hc.complete, 3)
```
Which states belong to which clusters
```{r}
# States in the first cluster
dat2[ind4.complete == 1,]
# States in the second cluster
dat2[ind4.complete == 2,]
# States in the third cluster
dat2[ind4.complete == 3,]
```

### (b) Hierarchically cluster the states using complete linkage and Euclidean distance, after scaling the variables to have standard deviation one.

```{r}
dat2_scale <- scale(dat2)
hc.complete_scale <- hclust(dist(dat2_scale), method = "complete")
```

Visualize the dendrogram and cut the dendrogram at a height that results in three distinct clusters

```{r}
fviz_dend(hc.complete_scale, k = 3,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)

ind4.complete_scale <- cutree(hc.complete_scale, 3)
```

Which states belong to which clusters
```{r}
# States in the first cluster
dat2_scale[ind4.complete_scale == 1,]
# States in the second cluster
dat2_scale[ind4.complete_scale == 2,]
# States in the third cluster
dat2_scale[ind4.complete_scale == 3,]
```

### (c) Does scaling the variables change the clustering results? Why? In your opinion, should the variables be scaled before the inter-observation dissimilarities are computed?

Yes, scaling the variables affects the clustering outcomes.

This is because when variables are on different scales, particularly in methods like hierarchical clustering that depend on distance calculations, variables with larger scales (eg. Assault) disproportionately influence the results.

I think the variables should be scaled before the inter-observation dissimilarities are computed to ensure that no single variable  influences the outcome purely because of its scale.